---
title: "Impact of food availability and light on S. purpuratus larval growth - data analysis"
output:
  word_document:
    keep_md: true
    fig_caption: true
  html_document: default
  pAl_df_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev = "tiff",
                      dpi = 300)
```



```{r loading libraries, message=FALSE, warning=FALSE}
library('dplyr')
library('readr')
library('ggplot2')
library('ggdist')
library('knitr')
library('tidyr')
library('emmeans')
library('rstan')
rstan::rstan_options(auto_write = TRUE)
library('brms')
options(mc.cores = parallel::detectCores())  # run all cores
library('bayesplot')
library('marginaleffects')
library('ggdist')
```


```{r}
nchain = 4
niter  = 2500
moment_match = TRUE
```


# 1. Data overview

Making a dataset containing only the data on __S. purpuratus__
```{r import data}
Sp_df <-  read_delim("larval_morphology.csv", delim = ",",
                  col_types = "fffnfiffffiniif")
Sp_df = drop_na(Sp_df, length)
# make Pl_df$larva by concatenating Pl_df$species with Pl_df$larva 
Sp_df$larva <- as.factor(paste0(Sp_df$species, Sp_df$larva))
Sp_df <- Sp_df[Sp_df$species == "Sp",]
Sp_df <- Sp_df[Sp_df$length > 0,]
Sp_df <- Sp_df[! is.na(Sp_df$length),]
#ensure correct order for levels
Sp_df<-Sp_df %>% mutate(lit = factor(lit, levels = c("DD", "LD", "LL")))
Sp_df<-Sp_df %>% mutate(rod = factor(rod, levels = c("BR", "PO", "ALA")))

print(paste0('There are ', dim(Sp_df)[1], ' measures from ', length(unique(Sp_df$larva)), ' individual larvae.'))
```

```{r}
head(Sp_df)
```

For the statistical analysis length needs to be scaled to L. 

```{r scaling variables}
meanL <- mean(Sp_df$length)
sdL <- sd(Sp_df$length)

Sp_df$L  <- as.numeric(scale(Sp_df$length))
Sp_df <- droplevels(Sp_df) # drop factor levels which are absent
head(Sp_df)
```


The chunk below produces a data summary for each condition. In column n we calculated also the number of observations.

```{r data summary}
Sp_df %>% group_by(species, dpf, lar_ml, lit, condition, fed) %>%
  summarise(mean = mean(length, na.rm = TRUE), stdev = sd(length, na.rm = TRUE),
            n = n())
```



## EXPERIMENTAL SETUP AND AIM

We are interested in investigating how the light-dark 
cycle (lit) influence the phenotypic response to food availability (fed): larvae grow much shorter arms when food is abundant to save maternal storages; on the contrary, when food is scarce they develop much longer arms to maximize their capability to collect food. This response is controlled by a dopaminergic signalling. To interfere with this signalling we used an in inhibitor of the TH (one of the enzymes involved in the dopamine synthesis) and DMSO as negative control. Three sets of spicules have been measured: Body Rod (BR), Post Oral (PO), and Anterolateral (ALA) arms .


## PLOT DATA OVERVIEW

```{r food impact, fig.height=5, fig.width= 7}
Sp_df %>%
  ggplot(aes(y=length,x=condition, fill=fed)) +
  facet_grid(lit~rod ) + 
    geom_boxplot(position = position_dodge(width = 0.75)) +
    geom_jitter(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), size=0.7)+ 
  scale_fill_manual(values=c("#00BBC1", "#F86D63"))+
 #geom_violin()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  ggtitle("S. purpuratus, dopamine and light control of phenotypic response")  
```

```{r light impact, fig.height=5, fig.width= 7}
Sp_df %>%
  ggplot(aes(y=length,x=condition, fill=lit, )) +
    facet_grid(fed~ rod) + 
    geom_boxplot(position = position_dodge(width = 0.75)) +
    geom_jitter(position = position_jitterdodge(jitter.width = 0.25, dodge.width = 0.75), size=0.7)+
    scale_fill_manual(values=c("#7473d1", "#f5b905","#d62222"))+
 #geom_violin()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  ggtitle("S. purpuratus, food and dopamine control on light response")  
```


# 2. Statistics


## PRIOR PREDICTIVE TEST


## M0 Single intercept

Set priors on slope. Weakly informative priors are used to guide the model.

Each treatment should be replicated sufficiently within each larva to capture the treatment effect accurately. Typically, having at least 3-5 measurements per treatment per larva can provide a reasonable balance between model complexity and data sufficiency.

Ideally, we would represent the data as a nested structure, with larvae nested within species. This would allow us to estimate the variance components for the species and larva levels. However, given the low number of measures per larva, it might be challenging to estimate these variance components reliably. Therefore, we will start with a simpler model that includes only the larva level as a random effect.

```{r}
get_prior(bf(L ~  1, sigma ~1), data=Sp_df)
```


```{r}
si_priors <- c(
  set_prior("normal(0, 0.5)", class = "Intercept"),  # prior for intercept
  set_prior("student_t(3, 0, 0.5)", class = "sd", group = "larva"), # Student's t prior for group-level sd
  set_prior("student_t(3, 0, 0.5)", dpar = "sigma", class = "Intercept")  # Student's t prior for residual sd
)
```


The zeroth model is built to check the effect of the group level and see how the data is distributed overall. The nesting structure explicitly acknowledges the hierarchical nature of the data. This helps to avoid pseudoreplication and ensures that the estimates of variance components are not biased.


```{r message=FALSE}
Sp_intercept_mod <- brm(
  bf(L ~  1 + (1|larva), sigma ~ 1), 
  family = gaussian,
  data=Sp_df, prior = si_priors,
  chains = nchain, 
  iter = niter, warmup = niter/2,
   save_pars = save_pars(all = TRUE)
  )
Sp_intercept_mod
```
```{r}
get_prior(bf(L ~  rod, sigma ~ rod), data=Sp_df)
```


```{r}
priors <- c(
  set_prior("normal(0, 1)", class = "Intercept"),  # prior for intercept
  set_prior("normal(0, 2)", class = "b"),
  set_prior("student_t(3, 0, 1)", class = "sd", group = "larva"),  # Student's t prior for group-level sd
  set_prior("student_t(3, 0, 1)", dpar = "sigma", class = "Intercept"),  # Student's t prior for residual sd
  
  set_prior("normal(0, 2)", class = "b", dpar = "sigma")
)
```



## M1 Rod model

```{r message=FALSE}
Sp_rod_mod <- brm(
  bf(L ~  condition*rod + (1|larva), sigma ~ condition*rod), 
  family = gaussian,
  data=Sp_df, prior = priors,
  chains = nchain,
  iter = niter, warmup = niter/2,
  save_pars = save_pars(all = TRUE)
  )
Sp_rod_mod
```



```{r}
plot(Sp_rod_mod, ask=FALSE, variable = "^b_", regex = TRUE)
```


## M2 Rod Fed model

```{r message=FALSE}
Sp_rod_fed_mod <- brm(
  bf(L ~  condition*rod*fed + (1|larva), sigma ~ condition*rod*fed), 
  family = gaussian,
  data=Sp_df, prior = priors,
  chains = nchain,
  iter = niter, warmup = niter/2,
   save_pars = save_pars(all = TRUE)
  )
Sp_rod_fed_mod
```


```{r}
plot(Sp_rod_fed_mod, ask=FALSE, variable = "^b_", regex = TRUE)
```



## M3 Rod Lit model

```{r message=FALSE}
Sp_rod_lit_mod <- brm(
  bf(L ~  condition*rod*lit + (1|larva), sigma ~ condition*rod*lit), 
  family = gaussian,
  data=Sp_df, prior = priors,
  chains = nchain,
  iter = niter, warmup = niter/2,
   save_pars = save_pars(all = TRUE)
  )
Sp_rod_lit_mod
```

```{r}
plot(Sp_rod_lit_mod, ask=FALSE, variable = "^b_", regex = TRUE)
```

## M4 condition Rod Fed Lit model

```{r Sp_rod_fed_lit_mod, message=FALSE}
Sp_rod_fed_lit_mod <- brm(
  bf(L ~  condition*rod*fed*lit + (1|larva), sigma ~ condition*rod*fed*lit), 
  family = gaussian,
  data=Sp_df, prior = priors,
  chains = nchain,
  iter = niter, warmup = niter/2,
   save_pars = save_pars(all = TRUE)
  )
Sp_rod_fed_lit_mod
```

```{r}
plot(Sp_rod_fed_lit_mod, ask=FALSE, variable = "^b_", regex = TRUE)
```

# 3. Models evaluation using LOO

```{r}
Sp_intercept_mod = add_criterion(Sp_intercept_mod,
                        criterion = "loo", moment_match = moment_match, reloo = TRUE)
Sp_rod_mod = add_criterion(Sp_rod_mod,
                        criterion = "loo", moment_match = moment_match, reloo = TRUE)
Sp_rod_lit_mod = add_criterion(Sp_rod_lit_mod,
                        criterion = "loo", moment_match = moment_match, reloo = TRUE)
Sp_rod_fed_mod = add_criterion(Sp_rod_fed_mod,
                        criterion = "loo", moment_match = moment_match, reloo = TRUE)
Sp_rod_fed_lit_mod = add_criterion(Sp_rod_fed_lit_mod,
                        criterion = "loo", moment_match = moment_match, reloo = TRUE)
```



```{r}
# Perform LOO comparison
loo_results <- loo_compare(Sp_intercept_mod,
                           Sp_rod_mod, 
                           Sp_rod_fed_mod,
                           Sp_rod_lit_mod,
                           Sp_rod_fed_lit_mod)
loo_results
```



```{r}
kable(loo_results)
```


```{r}
best_model_name <- rownames(loo_results)[1]
best_model <- get(best_model_name)

# Save the best model to an RDS file
saveRDS(best_model, file = paste0(
  "./model_objects/", best_model_name, ".rds"))

# Print the name of the best model
print(paste("The best model is:", best_model_name))
```


```{r}
#saveRDS(best_model, file = paste0("./model_objects/", best_model_name, ".rds"))
```

